Honor Among Thieves^[I thank Vlad Tarko and Elana Safner for helpful comments. All remaining errors are my own.]

### 2022-12-22 response_to_reviewers.qmd

---
title: "Response to Reviewers"
format: pdf
editor: visual
---

Dear Reviewers,\
\
Thank you very much for reviewing and providing comments on my manuscript, they have been very helpful in improving it. In addition to grammatical fixes and minor edits to improve clarity, I have made two major changes to the paper in response to your suggestions.\
\
First, I have added a short discussion at the end of section 2.1 to respond to Reviewer 1's request for examples of uncommitted publishers. It seems to have been that priority rights were respected without costly commitments (paying for advance sheets, negotiating permission from the author, etc) in cases of new & unknown authors. Here, exclusivity was not seen as too valuable as there was little competition to publish these authors. These are primarily for publishers to undertake experiments, and only if the gamble pays off and the author becomes popular enough to keep reprinting, the firm that first published them still has rights to exclusivity under trade courtesy.

Second, I have added a discussion to expand upon the role of reputation and repeated dealings constraining defection, using the example of a simple repeated Prisoners' Dilemma and trigger strategies, as suggested by the editor. I discuss a firm's reputation as upholding trade courtesy and avoiding piracy as a depreciable capital asset that a publisher risks devaluing if they turn to piracy.

In response to Reviewer 1's question about lesser sanctions (examples of compensation) that prevented piracy - I do provide a few examples of how disputes were settled among followers of courtesy. It is certainly difficult to deter a dedicated pirate publisher with anything but a price war. But for the "legitimate" publishers that occasionally stepped on each other's toes, I provide several examples in that section, as well as section 2.6, about how publishers resolved such disputes with "adjustments" or in-kind bargains where the "rights" are often shared among publishers in different forms - the right to paperback editions vs. hardback, serials, present vs. future editions. Admittedly some of this was a structural decision about where to put these examples in the paper to have sufficient material for each of the 8 design principles.\
\
Thank you again very much for your comments and for the opportunity to submit this paper.

### 2022-12-22 manuscript.Rmd
ALER, and add "commons" to keywords
Edits for editor

abstract: |
  | From 1790 to 1891, the United States prevented foreign authors from obtaining domestic copyright protection, implicitly subsidizing a domestic reprinting industry. With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce pseudo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection. This paper analyzes this system using the Bloomington School's institutional design principles to understand its effectiveness and pitfalls in managing the commons of unprotected foreign works in 19^th^ Century America.
  |
  | *JEL Classification:* H41, K29, N40, O34
  | *Keywords:* Copyright, intellectual property, literary piracy, publishing, economic history

@@

> "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience ... At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." - Henry Holt, quoted in @Gilbert1993 [31]

From 1790 to 1891, the United States explicitly prevented foreign authors from obtaining domestic copyright protection for their works.[^1] This policy, in part, was a boon to the domestic publishing industry, which largely reprinted books by foreign authors for an American audience [@Safnerpiratethyneighbor]. American publishers at the time, deemed "pirates" by their detractors, treated unprotected foreign works as a "free resource" held in common. As American publishers began to "print on" each other (by reprinting each other's reprints of foreign works), they were forced to develop a system to prevent the foreign-work commons from being depleted by ruinous competition and price wars. This voluntary system of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid- to the late-19^th^ century, when the U.S. began extending copyright protection to foreign authors.

[^1]: Even after the 1891 Chace Act that first gave copyright protection in the U.S. to foreign authors, it was *de facto* quite costly for authors to attain this. Additional requirements, some overtly protectionist such as the "manufacturing clauses" requiring foreign works to be printed in the U.S. by U.S. printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention, which standardized international copyright, until 1989.

This paper examines the effectiveness of the courtesy of the trade system as a strategy by publishers to manage the common pool resource of unprotected foreign works. I use the Bloomington School's "design principles" for robust institutions of governance [@Ostrom1990; @Ostrom2005; @Wilson2013; @Aligica2013; @Tarko2016, Ch.4] to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulate the procedures and protections that actual copyright protection would have legally provided. While it was not uniformly followed by all publishers, saw occasional breakdown into price wars, and ultimately faded with the formal recognition of foreign copyrights, trade courtesy serves as an interesting example of both how informal norms can resolve disputes, as well as an alternative mechanism of protecting intellectual property without statutory law.

Economists typically define a commons, or a common pool resource, as being rivalrous but non-excludable: each person's consumption subtracts from another's ability to consume, but all such users are unable to prevent others from accessing the resource. @Hardin1968 famously termed the situation a "tragedy of the commons," as the inability of users to exclude others' use logically implies users will rapidly deplete the resource. A classic solution, among economists, has been to establish property rights over the resource, allowing owners to exclude other users, internalizing the benefits and responsibility to improve and stewarding over the resource [@Alchian1965; @Demsetz1964]. However, establishing and enforcing property rights is costly, and is only done when the net benefits of "propertization" exceed the net costs of maintenance (or the net benefits of a commons) [@Demsetz1969]. Indeed, economists have shown that if rights are designed poorly and allocated from above, then a symmetric tragedy of the *anti*-commons may result, where any one possessor of exclusion-rights among many can veto the productive use of a resource [@Heller1997; @Buchanan2000; @Heller2008].

@@

@DouradoTabarrok2015 call this the "eightfold path to success" and apply this to ideas and inventions, which they see as a "Super-Lockean commons" which, if accessible, actually *grows* the more that people productively use it (c.f. @Hess2007 [5]). @Safner2016 applies these 8 principles to explain the success of Wikipedia, the free online encyclopedia, in managing public knowledge.

Federally-sponsored American literary piracy has been noted and extensively discussed by lawyers, historians, and literary scholars. Many have examined the legal development of American (and previously, British) copyright laws, noting the lapse in protection to foreign authors in the United States throughout the 19^th^ century (among the most systematic are @Patterson1968 and @Johns2009). A number have also examined the movement for international copyright in the United States that achieved partial victory at the end of the 1880s (see, e.g., @Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).

While there is a vast literature in the economics of copyright and intellectual property more broadly,[^2] economists have largely overlooked the dynamics of this period of "piracy" except in the service of broader arguments or economic history regarding intellectual property rights. @Khan2005, however, recognizes the idiosyncratic and endogenous component of copyright history, primarily to distinguish the unique and superior American system of IP from that of European systems. @Khan2004 examines its welfare effects on American authors, publishers, and consumers, finding it a net positive. @Spoo2013 [Ch.1] notably places the workings of the courtesy system in the literature of solutions to the tragedy of the commons, citing touchstone works such as @Ellickson1994 on neighbors solving disputes and @Rustiala2012 on norms in fashion, comedy, and cooking. However, the system has yet to be explicitly analyzed in a fuller framework to explain its origin, mechanisms accounting for its success, and decline, as an institution. This paper examines how the set of norms and practices constituting "trade courtesy" in the 19<sup>th</sup> century adhere to the 8 principles of success outlined above, as well as how they led to the emergence and eventual decline of this system of self-regulation.

[^2]: See e.g. @Besen1991; @Posner2003 for a good survey. For a growing corpus of more critical works, see e.g. @Boldrin2008; @Kinsella2008; @Heller2008; @Rustiala2012; @Bell2014; @DouradoTabarrok2015

@@

> "[Trade courtesy] was a brief realization of the ideals of philosophical anarchism--- self-regulation without law. ---Henry Holt (1908)2

Statutory copyright came to the United States with the Copyright Act of 1790, its authority derived from the Copyrights Clause of the U.S. Constitution[^3] [@Pattersonbook]. Largely mimicking the 1710 Statute of Anne in Britain - the first modern copyright statute - it granted "sole right and liberty of printing, reprinting, publishing and vending" to authors of "any map, chart, book or books already printed within these United States" for "the term of fourteen years from the recording the title thereof in the clerk's office," with an optional 14-year renewal term (Copyright Act of 1790, \S 1). Crucially for American publishers, section 5 of the Act explicitly denies copyright protection to foreigners:

[^3]: "To promote the Progress of Science and useful Arts, by securing for limited Times to Authors and Inventors the exclusive Right to their respective Writings and Discoveries" (U.S. Constitution, Article I, \S8, Cl. 8).

@@

The domestic publishing industry in the early United States was largely run by Scottish and Irish immigrants such as Mathew Carey, who transplanted their tried-and-true experiences pirating London's books in Edinburgh and Dublin into New York and Philadelphia [@Johns2009, p.188ff; @Spoo2013, 34; @Groves2007b, 141].[^4] Early politicians and pundits saw reprinting as key to the emerging "American system" of political economy championed by Alexander Hamilton, Henry Clay, and Friedrich List, whereby tariffs and other measures would protect infant manufacturing industries. Thinkers of this school saw the reprinting (authorized or not) of European (predominantly British) works as key to instilling mass literacy, inspiration for innovation, instilling republican virtue [@Johns2009; @Spoo2013, 28; @Clark1960, 70]. In the first few decades of the American colonies and early republic, the great distances between major cities kept parochial printing presses from cannibalizing each other's hold over their local market.

@@

At mid-century, there were about 400 publishers concentrated in New York, Philadelphia, Baltimore, and Boston, according to the authoritative *Trubner's Bibliographical Guide to American Literature*, compiled in 1859, [@Shove1936, 23]. Harper & Brothers was the largest firm, with over 2 million volumes sold in 1859.

The growth of the industry brought about fierce competition, especially after the Depression in the late 1830s and a ruinous price war in the 1840s. The Philadelphia publisher Carey & Lea, for example, saw their output of new titles shrink by over 50% between 1835 and 1837 [@Groves2007b, 141-142]. Upstart periodicals such as *Brother Jonathan* and *New World* responded to an uptick in demand for literature by resorting to clever alternative forms of publishing: "monthlies" and "weeklies" printed entire English novels in small, low quality type on "mammoth" size single sheets sent through the mail direct to customers in newspaper format, exploiting cheap postal rates [@Groves2007b, 142; @Barnes1974, 4-24; @Madison1966,24-25]. This heavily depressed book prices, forcing publishers to respond in kind (as did Harper & Brothers), or to exit the market (like Wiley and Blanchard & Lea, both of whom focused on copyrightable textbooks) [@Madison1966,25][^5]. This period of "cutthroat competition" caused publishers to recognize the need for self-regulation and organizational solutions within the industry into the 1840s. Some publishers, such as Carey, suggested forming an explicit organization modeled off of the London Stationers Company. Carey set up The Philadelphia Company as an explicit cartel arrangement, only to see it evaporate 5 years later.[^6] Despite its failure, @Johns2009 [199] summarizes the common sentiment within the publishing industry:

@@

[^6]: Once such a cartel allocated titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint. Without an inflexible cartel, a bargain could be reached between two publishers to establish boundaries, but this was not possible with a large organization and rogue outside printers.

@@

```{r, echo =F, message=F, warning=F}
library(tidyverse)
tribble(
  ~"Date/s", ~"Event",
  "1790", "Copyright Act exempts foreign authors from copyright protection in U.S.",
  "1800-1801", "Acts of Union establishes copyright protection of British authors in Ireland",
  "1816", "First tariff on imported books",
  "c.1830s-1840s", "First price war",
  "1850s", "More explicit formulation of courtesy",
  "c.1870s-1880s", "Second price war",
  "1891", "Copyright Act grants foreign authors copyright protection in U.S."
) %>%
  knitr::kable(format = "latex",
               caption = "Outline of Courtesy Period in United States",
               booktabs = TRUE)
```

@@

The foreign reprinting industry in the United States was spawned directly from Section 5 of the Copyright Act of 1790. An American publisher could publish works by *American* authors only by acquiring the copyright from them first. Since section 5 exempted *foreign* authors from copyright protection, it provided an indirect subsidy to reprint these works without concern for acquiring valid copyrights. As such, rather than paying and haggling with an *American* author for her copyright, an American publisher could print European authors at a significantly reduced cost. Both established and upstart publishers rapidly entered this market, searching for a quick opportunity to republish a "sure thing" to an American populace hungry for English literature.

The informal "courtesy of the trade" system of norms was both voluntary and ever-changing with market conditions. Not all publishers abided by trade courtesy, and some that followed norms closely for one decade would later violate them when it was to their benefit. Some firms that had cut their teeth in the profession by reprinting other publishers' reprints of foreign works later turned "legitimate" by following trade courtesy and enforcing it against upstart pirates.

@@

By extension of the existing principles, publishers following trade courtesy also developed an "options" system for firms that had previously published a specific author's work, whereby other firms would generally recognize that publisher as being "associated" with the author, and therefore obtain first claim on new works by that author. This was referred to in the trade by the "rule of association" [@Spoo2013,38]. Aside from providing a bright line rule for priority of claims, a major benefit of the association rule was the a firm could advertise themselves as being the "authorized" publisher of a foreign author [@Spoo2013,42].

However, simply asserting that one *claimed* first priority in publishing an author was insufficient, and could incentivize "courtesy creep," [@Spoo2013,37]: Publishers might announce simultaneously or simply announce their intent to print any title that sounded promising, hoping to establish priority even if they would never print. Such a "first to announce" rule could quickly lead to inefficient rent-seeking: claimants merely assert claim in order to block others from using it, even if the claimant never intends to develop the resources in a productive way [@Posner2003,17]. Such a rule risks creating an "anti-commons" where multiple firms would seek to block any one firm from production [@Heller1997; @Heller1998; @Heller2008].

Instead of a "first to announce rule," a "committed-searcher rule," where the first claimant to undertake costly steps (strongly correlated with actually developing the resource) improves upon these incentives [@Posner2003,17]. Publishers following trade courtesy required a firm seeking to stake a claim to undertake a more costly action, often reaching a public agreement with the foreign author, or more frequently, purchasing advance sheets of the edition abroad. Holt summarizes: "If a publisher had the advance sheets in his possession, such right or claim overrode a simple announcement," (quoted in @Harper1912, 110-11). @Spoo2013 [37] describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown, and the firm undertaking a risky experiment.

@@

To sustainably manage a commons, users must obtain benefits proportionate to the costs they bear. Otherwise, if a substantial subset of the group free rides without contributing to the commons, or contributes without any reward, the tragedy of the commons may rear its ugly head, and the resource will be depleted. In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs.

@@

Naturally, upstart publishers that were only in it for a quick buck had little incentive to worry about reputation. So long as there were not too many publishers competing over price in this way, the industry could maintain this equilibrium.

Upholding proper "gentlemanly" conduct was not costless, it did require publishers to sacrifice profitable opportunities in the hope of not destroying their reputation (and hopes of future profit opportunities). As mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher *A* seeking to reprint works of foreign author *X* (and hoping to beat publisher *B* to the punch) could not merely assert this and hope publisher *B* would back down; publisher *A* needs to strike an exclusive arrangement with author *X* or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.

This required investment led to one of the great ironies of the system: despite the absence of a legal duty, U.S. publishers frequently paid significant honoraria to British authors for the priority "right" to print their works. In fact, publishers often *emphasized* in their advertisements the amount they paid to foreigners for their "exclusive" or "authorized" edition [@Spoo2013, 24]. Occasionally, it would be payments to the British publisher for advance sheets. Harper & Brothers, for example, paid Edward Bulwer-Lytton £50 per volume and Charles Dickens between £250 and £2,000 for his novels [@Spoo2013, 39]. Harper Brothers, the largest 19<sup> th</sup> century publisher in the U.S. (and later the world) described the courtesy practice of purchasing advance sheets from foreign authors or their agents or publishers:

> "In the absence of an international copyright, it is the custom for an English author, or his agent in London, to send early sheets to some American publisher, fixing a price therefor, and by a law of courtesy the American publisher who has issued the previous works of an author is entitled to the first consideration of that author's new book," [@Harper1912,358].

This practice of paying for advance sheets gave an aura of semi-legitimacy to the courtesy system. Publisher Henry Holt, in his 1893 reflections on trade courtesy, notes that "It not only prevented ruinous competition between American publishers, but also secured to foreign authors most of their rights," (ibid, 32). While many popular British authors publicly lamented the lack of copyright protection in America --- notably Charles Dickens, who made it a feature of his 1867 tour of the U.S. --- they were often compensated even though publishers had no obligation to do so. Such honoraria and payment for the "exclusivity" that came with being first, or with the author's "authorized" version, was surely less than the author would have received in royalties or profit-sharing under copyright protection. British author Frederick Marryat estimated in 1839 that if Walter Scott (the British author who first became a best seller in the U.S.) lost out on \$500,000 in royalties from U.S. sales [@Madison1966,8-12]. Carey & Lea, his "authorized" American publisher made "token" payments (by comparison) --- £50 for Canongate in 1824, £250 for Life of Napoleon in 1827 (which was the largest such U.S. payment to that date) [@Madison1966,12]. To further underscore the copyright tradeoff between U.S. and Britain, Scott's novels often sold for 1/4 of the British price in the U.S., but gave him more American readers than English readers in total (ibid,12).

>>>>

abstract: |
  | From 1790 to 1891, the United States prevented foreign authors from obtaining domestic copyright protection, implicitly subsidizing a domestic reprinting industry. With foreign works a "free" and unprotected resource, American publishers created a system of voluntary norms, known as "trade courtesy" to create and enforce pseudo-property rights in uncopyrighted foreign works, simulating the effects of legal copyright protection. This paper analyzes this system using the Bloomington School's institutional design principles to understand its effectiveness and pitfalls in managing the commons of unprotected foreign works in 19^th^ Century America.
  |
  | *JEL Classification:* H41, K29, N40, O34
  | *Keywords:* Copyright, intellectual property, literary piracy, publishing, economic history, commons

@@

> "[Publishing in 19^th^ Century America was] perhaps the greatest paradox in human experience ... At one end, its principal material was not protected by law, and the business lived to a large extent on what was morally, if not legally, thievery; while at the other end, there was honor among thieves, in the respect they paid each other's property." — Henry Holt, quoted in @Gilbert1993 [31]

From 1790 to 1891, the United States explicitly prevented foreign authors from obtaining domestic copyright protection for their works.[^1] This policy, in part, was a boon to the domestic publishing industry, which largely reprinted books by foreign authors for an American audience [@Safnerpiratethyneighbor]. American publishers at the time — deemed "pirates" by their detractors — treated unprotected foreign works as a "free resource" held in common. As American publishers began to "print on" each other (the industry's term for reprinting each other's reprints of foreign works without permission), they were forced to develop a system to prevent the foreign-work commons from being depleted by ruinous competition and price wars. This voluntary system of informal norms, known as "courtesy of the trade" among publishers, flourished from the mid- to the late-19^th^ century, when the U.S. began extending copyright protection to foreign authors.

[^1]: Even after the 1891 International Copyright Act first gave copyright protection in the U.S. to foreign authors, it was *de facto* quite costly for authors to attain this. Additional requirements, some overtly protectionist such as the "manufacturing clauses" requiring foreign works to be printed in the U.S. by U.S. printers, frequently frustrated foreign authors and led to the failure of the United States to qualify for the Berne Convention, which standardized international copyright recognition, until 1989.

This paper examines the effectiveness of the courtesy of the trade system as a strategy by publishers to manage the common pool resource of unprotected foreign works. I use the Bloomington School's "design principles" for robust institutions of governance [@Ostrom1990; @Ostrom2005; @Wilson2013; @Aligica2013; @Tarko2016, Ch.4] to explore how the system was largely successful in creating and enforcing pseudo-property rights that largely simulated the procedures and protections that actual copyright protection would have legally provided. While it was not uniformly followed by all publishers, saw occasional breakdown into price wars, and ultimately faded with the formal recognition of foreign copyrights, trade courtesy serves as an interesting example of both how informal norms can resolve disputes, as well as an alternative mechanism of protecting intellectual property without statutory law.

Economists typically define a commons, or a common pool resource, as being rivalrous but non-excludable: each person's consumption subtracts from another's ability to consume, but all such users are unable to prevent others from accessing the resource. @Hardin1968 famously termed the situation a "tragedy of the commons," as the inability of users to exclude others' use logically implies users will have an individual incentive to overuse the resource, leading to its depletion. A classic solution suggested by economists is to establish property rights over the resource which allows owners to exclude other users, internalizing the benefits and responsibility to improve and stewarding over the resource [@Alchian1965; @Demsetz1964]. However, establishing and enforcing property rights is costly itself, and is only done when the net benefits of "propertization" exceed the net costs of establishing and enforcing property (or the net benefits of a commons) [@Demsetz1969]. Indeed, economists have shown that if rights are designed poorly and allocated from above, then a symmetric tragedy of the *anti*-commons may result, where any one possessor of exclusion-rights among many can veto the productive use of a resource [@Heller1997; @Buchanan2000; @Heller2008].

@@

@DouradoTabarrok2015 call this the "eightfold path to success" and apply this to ideas and inventions, which they see as a "Super-Lockean commons" which, if accessible, actually *grows* the more that people productively use it (c.f. @Hess2007 [5]). @Safner2016wiki applies these 8 principles to explain the success of Wikipedia, the free online encyclopedia, in managing public knowledge.

State-sponsored American literary piracy has been noted and extensively discussed by lawyers, historians, and literary scholars. Many researchers have examined the legal development of American (and previously, British) copyright laws, noting the lapse in protection to foreign authors in the United States throughout the 19^th^ century (among the most systematic are @Patterson1968 and @Johns2009). A number have also examined the movement for international copyright recognition in the United States that achieved partial victory at the end of the 1880s (see, e.g., @Putnam1891; @Kampelman1947; @Dozer1949; @Henn1953; @Clark1960; @Ringer1968; @Barnes1974).

While there is a vast literature in the economics of copyright and intellectual property more broadly,[^2] economists have largely overlooked the dynamics of this period of "piracy" except in the service of broader arguments or economic history regarding intellectual property rights. Economists primarily focus on the public good aspects of ideas and expressions: once published, an idea can be utilized by many simultaneously, and that it becomes difficult for the initial producer to appropriate the value of their idea by excluding others (competitors and consumers) from using the idea without payment. One of the traditional economic arguments for copyright protection is that it makes such ideas, instantiated in books or other media, excludable — competitors and consumers risk legal liability for using protected works without the initial producer's consent. As such, publishers traditionally negotiate with authors to acquire copyright to their works, to enforce against rivals. @Khan2005 recognizes the idiosyncratic and endogenous nature of copyright history, including the period of U.S. literary piracy, primarily to distinguish the unique and superior American system of IP from that of European systems. @Khan2004 examines the welfare effects on American authors, publishers, and consumers of the systematic pirating of foreign works, finding it a net positive. @Spoo2013 [Ch.1] notably places the workings of the courtesy system in the literature of solutions to the tragedy of the commons, citing touchstone works such as @Ellickson1994 on neighbors solving disputes and @Rustiala2012 on norms in fashion, comedy, and cooking. However, the system has yet to be explicitly analyzed in a fuller framework to explain its origin, mechanisms accounting for its success, and decline, as an institution. This paper examines how the set of norms and practices constituting "trade courtesy" in the 19<sup>th</sup> century adhere to the 8 principles of success outlined above, as well as how they led to the emergence and eventual decline of this system of self-regulation.

[^2]: See e.g. @Besen1991; @Posner2003 for good surveys of the economic issues of intellectual property. For a growing corpus of more critical works, see e.g. @Boldrin2008; @Kinsella2008; @Heller2008; @Rustiala2012; @Bell2014; @DouradoTabarrok2015.

@@

> "[Trade courtesy] was a brief realization of the ideals of philosophical anarchism--- self-regulation without law. ---Henry Holt (1908, 2)

Statutory copyright came to the United States with the Copyright Act of 1790, its authority derived from the Copyrights Clause of the U.S. Constitution[^3] [@Pattersonbook]. Largely mimicking the 1710 Statute of Anne in Britain — the first modern copyright statute — it granted "sole right and liberty of printing, reprinting, publishing and vending" to authors of "any map, chart, book or books already printed within these United States" for "the term of fourteen years from the recording the title thereof in the clerk's office," with an optional 14-year renewal term (Copyright Act of 1790, \S 1). Crucially for American publishers, section 5 of the Act explicitly denies copyright protection to foreigners:

[^3]: "The Congress shall have Power...To promote the Progress of Science and useful Arts, by securing for limited Times to Authors and Inventors the exclusive Right to their respective Writings and Discoveries" (U.S. Constitution, Article I, \S8, Cl. 8).

@@

The domestic publishing industry in the early United States was largely run by Scottish and Irish immigrants such as Mathew Carey, who transplanted their tried-and-true experiences pirating London's books in Edinburgh and Dublin to New York and Philadelphia [@Johns2009, p.188ff; @Spoo2013, 34; @Groves2007b, 141].[^4] Early politicians and pundits saw reprinting as key to the emerging "American system" of political economy championed by Alexander Hamilton, Henry Clay, and Friedrich List, whereby tariffs and other measures would protect infant manufacturing industries. Such advocates saw the reprinting (whether authorized or not) of European (predominantly British) works as key to promoting mass literacy, inspiration for innovation, and instilling republican virtue [@Johns2009; @Spoo2013, 28; @Clark1960, 70]. In the first few decades of the American colonies and early republic, the great distances between major cities kept parochial printing presses from cannibalizing each other's hold over their local market. New York publishers faced little competitive threat from Philadelphia or Boston-based publishing houses, and vice versa.

@@

At mid-century, there were about 400 publishers concentrated in New York, Philadelphia, Baltimore, and Boston, according to the authoritative *Trubner's Bibliographical Guide to American Literature*, compiled in 1859, [@Shove1936, 23]. Harper & Brothers was the largest firm, with over 2 million volumes sold in 1859. The growth of the industry brought about fierce competition, especially after the Depression in the late 1830s and a ruinous price war in the 1840s. The Philadelphia publisher Carey & Lea, for example, saw their output of new titles shrink by over 50% between 1835 and 1837 [@Groves2007b, 141-142]. Upstart periodicals such as *Brother Jonathan* and *New World* responded to an uptick in demand for literature by resorting to clever alternative forms of publishing: "monthlies" and "weeklies" printed entire English novels in small, low quality type on "mammoth" size single sheets sent through the mail direct to customers in newspaper format, exploiting cheap postal rates [@Groves2007b, 142; @Barnes1974, 4-24; @Madison1966,24-25]. This heavily depressed book prices, forcing publishers to respond in kind (as did Harper & Brothers), or to exit the market (like Wiley and Blanchard & Lea, both of whom focused on copyrightable textbooks) [@Madison1966,25][^5]. This period of "cutthroat competition" caused publishers to recognize the need for self-regulation and organizational solutions within the industry into the 1840s. Some publishers, such as Carey, suggested forming an explicit organization modeled off of the London Stationers Company, the state-sponsored guild that controlled and regulated publishing in Britain since the 16^th^ century (discussed below in Section 2.7). Carey set up The Philadelphia Company as an explicit cartel arrangement, only to see it evaporate 5 years later.[^6] Despite its failure, @Johns2009 [199] summarizes the common sentiment within the publishing industry:

@@

[^6]: Once such a cartel created a public Register to allocate exclusive titles to its members to print, and prevented other internal members from printing on each other, outside publishers not part of the arrangement could observe which titles were "a sure thing" to reprint. Without an inflexible cartel, a bargain could be reached between two publishers to establish boundaries, as described below, but this possibility would be hindered by a large organization and rogue outside printers.

@@

```{r, echo =F, message=F, warning=F}
library(tidyverse)
tribble(
  ~"Date/s", ~"Event",
  "1790", "Copyright Act exempts foreign authors from copyright protection in U.S.",
  "1800-1801", "Acts of Union establishes copyright protection of British authors in Ireland",
  "1816", "First tariff on imported books",
  "c.1830s-1840s", "First price war",
  "1850s", "More explicit formulation of courtesy",
  "c.1870s-1880s", "Second price war",
  "1891", "International Copyright Act grants foreign authors copyright protection in U.S."
) %>%
  knitr::kable(format = "latex",
               caption = "Outline of Courtesy Period in United States",
               booktabs = TRUE)
```

@@

The foreign reprinting industry in the United States was spawned directly from Section 5 of the Copyright Act of 1790. An American publisher could publish works by *American* authors only by negotiating for the copyright from them first. Since section 5 exempted *foreign* authors from copyright protection, it provided an indirect subsidy to reprint these works without concern for acquiring valid copyrights. As such, rather than paying and haggling with an *American* author for her copyright, an American publisher could, in principle, print European authors at a significantly reduced cost. Both established and upstart publishers rapidly entered this market, searching for a quick opportunity to republish a "sure thing" to an American populace hungry for literature.

The informal "courtesy of the trade" system of norms was both voluntary and ever-changing with market conditions. Not all publishers abided by trade courtesy, and some that upheld the norms closely for a decade would later violate them when it was to their benefit. Likewise, some firms that had cut their teeth in the profession by reprinting other publishers' reprints of foreign works later turned "legitimate" by following trade courtesy and enforcing it against new upstart pirates.

@@

By extension of the existing principles, publishers following trade courtesy also developed an "options" system for firms that had previously published a specific author's work, whereby other firms would generally recognize that publisher as being "associated" with the author, and therefore obtain first claim on new works by that author. This was referred to in the trade by the "rule of association" [@Spoo2013,38]. Aside from providing a bright line rule for priority of claims, a major benefit of the association rule was that a firm could advertise themselves as being the "authorized" publisher of a foreign author [@Spoo2013,42].

However, simply asserting that one *claimed* first priority in publishing an author was insufficient, and could incentivize "courtesy creep," [@Spoo2013,37]: Publishers might announce simultaneously or simply announce their intent to print *any* title that sounded promising, hoping to establish priority even if they would never print it. Such a "first to announce" rule could quickly lead to inefficient rent-seeking: claimants merely assert claim in order to block others from using it, even if the claimant never intends to develop the resources in a productive way [@Posner2003,17]. Such a rule risks creating an "anti-commons" where multiple firms would seek to block any one firm from production [@Heller1997; @Heller1998; @Heller2008].

Instead of a "first to announce rule," a "committed-publisher rule," where the first claimant to publicly undertake costly steps (strongly correlated with actually developing the resource) improves upon these incentives [@Posner2003,17]. Publishers following trade courtesy required a firm seeking to stake a claim to undertake a more costly action, often reaching a public agreement with the foreign author, or more frequently, purchasing advance sheets of the edition abroad. Holt summarizes: "If a publisher had the advance sheets in his possession, such right or claim overrode a simple announcement," (quoted in @Harper1912, 110-11). @Spoo2013 [37] describes the rather intricate precision necessary for such a claim: purchasing copies abroad and merely importing them does not count as substantial action towards a claim (since it required no American labor), and mere announcement was only sufficient in cases where foreign authors were unknown. Holt describes that these cases were “a field for experiment, and if a publisher concluded one was worth experimenting with, though not worth paying for in advance of experiment, the rights from first announcement were intended to secure him the fruits of his experiment, if successful,” [@Publisherslibel, 360]. In such cases of an unknown or new author, "property rights" over the exclusivity were relatively low in value, hence costly investments in securing exclusivity, such as purchasing advance sheets, were not required due to lack of competition. However, if the gamble of publishing a new author paid off and was a hit, the exclusive right to publish becomes more valuable, and that publisher had established priority through their actions.

@@

To sustainably manage a commons, users must obtain benefits proportionate to the costs they incur. Otherwise, if a substantial subset of the group free rides without contributing to the commons, or contributes without any reward, the tragedy of the commons may rear its ugly head, and the resource will be depleted. In the case of the trade courtesy system, the constant existential threat was that publishers would print on each other and drive the price of works below production costs, making it unprofitable for *anyone* to publish foreign uncopyrighted works.

@@

The logic of repeated dealings often was sufficient to constrain publishers from violating these norms. Consider a simple game where publisher $A$ and publisher $B$ (both of whom republish foreign authors in the United States), can *cooperate* with one another by upholding trade courtesy, or they can *defect* and print on each other. If both publishers uphold trade courtesy and cooperate, they each earn payoff $b$. If one cooperates and another defects, the cooperator earns payoff $d$ and the defector earns payoff $a$. If both defect, they each earn payoff $c$. In this simple prisoners’ dilemma, payoffs $a>b>c>d$. The game is summarized in Table 2.

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(kableExtra)
game <- tribble(
  ~Player, ~Choice, ~Cooperate, ~Defect,
  "Publisher A", "Cooperate", "b, b", "d, a",
  "Publisher A", "Defect",  "a, d", "c, c"
)
game %>% 
  mutate(Player = text_spec(Player, format = "latex")) %>%
  mutate(Choice = text_spec(Choice, format = "latex")) %>% 
  rename(` ` = Player, `  ` = Choice) %>% 
  kbl(escape = FALSE, align = c("c", "c", "c", "c"), caption = "Payoffs a > b > c > d") %>% 
  kable_styling(bootstrap_options = "basic", full_width = FALSE, position = "center") %>% 
  collapse_rows(1) %>% 
  add_header_above(c(" " = 2, "Publisher B" = 2), escape = FALSE)
```

In a one-shot game, each publisher has a dominant strategy to defect and pirate the work of other domestic publishers. Thus, both publishers earn $c$ in the Nash equilibrium. If the publishers could find a way to ensure that both cooperate, they would both be better off, each earning $b$. Trade courtesy and the attendant series of norms served primarily as this coordination device. Specifically, the logic of repeated dealings and the reputation that individual publishers cultivated as participants in this system constrained the incentive to defect. If this game is repeated, assuming publishers have a common future discount factor $\delta$, the folk theorem famously says that many equilibria are possible where both publishers cooperate (see e.g. @Axelrod1984; @DixitSkeath). Considering the limiting case of the "grim trigger strategy," where each publisher cooperates until the other defects, and then defects for all future turns, the threat of retaliation constrains individual publishers to cooperate so long as:

$$\frac{b}{1-\delta}>a + \frac{\delta c}{1-\delta}$$
i.e. so long as the discount factor is sufficiently high, $\delta > \frac{b-a}{c-a}$. Reputation serves as a depreciable capital asset, held hostage for a publisher's good behavior as a participant in this "gentlemanly" system of norms [@KleinAlchian; @Williamsonhostages]. Once one becomes known as a "pirate," others will no longer respect claims to priority or association with any author.

Naturally, upstart publishers that were only in it for a quick buck and then close up shop had little incentive to worry about reputation. So long as there were not too many publishers competing over price in this way, the industry could maintain this equilibrium.

Upholding proper "gentlemanly" conduct was not costless, it did require publishers to sacrifice profitable opportunities in the hope of not destroying their reputation (and hopes of future profit opportunities). As first mentioned above, the rule of association raised the private marginal cost of an individual publisher trying to reprint foreign materials to avoid devaluing the resource of potential republications: Publisher *A* seeking to reprint works of foreign author *X* (and hoping to beat publisher *B* to the punch) could not merely assert this and hope publisher *B* would back down; publisher *A* needs to strike an exclusive arrangement with author *X* or purchase their sheets --- either way, incurring a higher marginal cost than other firms --- in order to do this.
